---
title: "Data Exploration on Heart Disease Data"
author: "Group 9"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
  word_document: default
---

# Univariate Exploration

# Load the Dataset
```{r}
# Load dataset

# Load necessary libraries
library(readr)
library(dplyr)  # %>% 

# Read the CSV file from Google Drive
url <- "https://drive.google.com/uc?export=download&id=12CriTyCML_9rGAySnmcitxXbij7N11T0"
my_data <- read_csv(url)

# View the first few rows
head(my_data)



```
```{r}
#Removing Unnecessary Columns

my_data <- my_data %>%
  select(-id, -dataset)
print(my_data)

```

```{r}
#Checking for Missing Data

my_data[!complete.cases(my_data),]
```

```{r}
#Handling Missing Values

my_data[is.na(my_data$trestbps),]


# Calculate the median
trestbps_medv <- median(my_data$trestbps, na.rm = TRUE)
print(trestbps_medv)

# Replace NA values with the median
my_data$trestbps[is.na(my_data$trestbps)] <- trestbps_medv

# Print updated data
print(my_data)


my_data[is.na(my_data$trestbps),]

```
```{r}
# Handling Missing Values

my_data[is.na(my_data$chol),]

chol_meanv<- mean(my_data$chol, na.rm=TRUE)
print(chol_meanv)

my_data$chol[is.na(my_data$chol)] <- chol_meanv

print(my_data)
my_data[is.na(my_data$chol),]
```
```{r}
# Handling Missing Values

my_data[is.na(my_data$thalch),]

thalch_meanv<- mean(my_data$thalch, na.rm=TRUE)
print(thalch_meanv)

my_data$thalch[is.na(my_data$thalch)] <- thalch_meanv

print(my_data)
my_data[is.na(my_data$thalch),]
```
```{r}
# Handling Missing Values

my_data[is.na(my_data$ca),]

# Find the most frequent value (mode) for 'ca'
mode_ca <- as.numeric(names(sort(table(my_data$ca), decreasing = TRUE))[1])
print(mode_ca)

my_data$ca[is.na(my_data$ca)] <- mode_ca


print(my_data)


my_data[is.na(my_data$ca),]

```

```{r}
# Handling Missing Values

my_data[is.na(my_data$thal),]

# Find the most frequent value (mode) for 'thal'
mode_thal <- as.character(names(sort(table(my_data$thal), decreasing = TRUE))[1])
print(mode_thal)


my_data$thal[is.na(my_data$thal)] <- mode_thal


print(my_data)


my_data[is.na(my_data$thal),]

```
```{r}
# Handling Missing Values

my_data[is.na(my_data$slope),]

# Find the most frequent value (mode) for 'slope'
mode_slope <- names(sort(table(my_data$slope), decreasing = TRUE))[1]
print(mode_slope)


my_data$slope[is.na(my_data$slope)] <- mode_slope


print(my_data)

my_data[is.na(my_data$slope),]

```




```{r}
#Checking for Remaining Missing Data

my_data[!complete.cases(my_data),]
```
```{r}
#Remove records with missing values
my_data<-my_data[complete.cases(my_data),]
print(my_data)
my_data[!complete.cases(my_data),]
```

# Summary Statistics
```{r}
# Summary of numeric variables
summary(my_data)

```
# Univariate Plots
```{r}

hist(my_data$age, 
     main="Histogram of Age", 
     col="lightblue", 
     xlab="Age", 
     border="black")

```
```{r}

hist(my_data$trestbps, 
     main="Histogram of trestbps ", 
     col="green", 
     xlab="trestbps", 
     border="black")

```
# Univariate Plots
```{r}

hist(my_data$chol, 
     main="Histogram of chol", 
     col="yellow", 
     xlab="chol", 
     border="black")

```


# Univariate Plots
```{r}

hist(my_data$thalch, 
     main="Histogram of thalch ", 
     col="blue", 
     xlab="thalch", 
     border="black")

```
```{r}

hist(my_data$oldpeak, 
     main="Histogram of oldpeak", 
     col="lightblue", 
     xlab="oldpeak", 
     border="black")

```
```{r}

hist(my_data$ca, 
     main="Histogram of Chest pain", 
     col="lightblue", 
     xlab="Ca", 
     border="black")

```
```{r}

hist(my_data$num, 
     main="Histogram of num", 
     col="lightblue", 
     xlab="num", 
     border="black")

```
```{r}
# Convert 'restecg' to numeric
my_data$restecg <- as.numeric(factor(my_data$restecg, levels = c("normal", "lv hypertrophy", "st-t abnormality")))

# Convert 'fbs' (fasting blood sugar) to numeric
my_data$fbs <- ifelse(my_data$fbs == TRUE, 1, 0)

# Convert 'slope' to numeric
my_data$slope <- as.numeric(factor(my_data$slope, levels = c("upsloping", "flat", "downsloping")))



my_data$num <- as.factor(my_data$num)

# Update the levels of 'num' to represent meaningful categories
levels(my_data$num) <- c("No Heart Disease", "Heart Disease -Type1", "Heart Disease -Type2", "Heart Disease -Type3","Heart Disease -Type4")


str(my_data)

# Print the first few rows to verify the new labels
head(my_data)

```
```{r}

hist(my_data$restecg, 
     main="Histogram of restecg", 
     col="lightblue", 
     xlab="restecg", 
     border="black")

```



```{r}

hist(my_data$fbs, 
     main="Histogram of fbs", 
     col="lightblue", 
     xlab="fbs", 
     border="black")

```


```{r}

hist(my_data$slope, 
     main="Histogram of slope", 
     col="lightblue", 
     xlab="slope", 
     border="black")

```




```{r}

# Density Plot (Smooth Distribution)
plot(density(my_data$age), 
     main="Density Plot of Age", 
     col="blue", 
     lwd=2)

```
```{r}

# Density Plot (Smooth Distribution)
plot(density(my_data$trestbps), 
     main="Density Plot of trestbps", 
     col="blue", 
     lwd=2)

```


```{r}

# Density Plot (Smooth Distribution)
plot(density(my_data$chol), 
     main="Density Plot of chol ", 
     col="blue", 
     lwd=2)

```


```{r}

# Density Plot (Smooth Distribution)
plot(density(my_data$thalch), 
     main="Density Plot of thalch", 
     col="blue", 
     lwd=2)

```

```{r}

# Density Plot (Smooth Distribution)
plot(density(my_data$oldpeak), 
     main="Density Plot of oldpeak", 
     col="blue", 
     lwd=2)

```

```{r}

# Density Plot (Smooth Distribution)
plot(density(my_data$slope), 
     main="Density Plot of slope", 
     col="blue", 
     lwd=2)

```


```{r}

# Density Plot (Smooth Distribution)
plot(density(my_data$restecg), 
     main="Density Plot of restecg", 
     col="blue", 
     lwd=2)

```



```{r}

# Boxplot 
boxplot(my_data$age, 
        main="Box Plot of Age", 
        col="orange")

```


```{r}

# Boxplot 
boxplot(my_data$trestbps, 
        main="Box Plot of trestbps ", 
        col="orange")

```
```{r}

# Boxplot 
boxplot(my_data$chol, 
        main="Box Plot of chol", 
        col="orange")

```

```{r}

# Boxplot 
boxplot(my_data$restecg, 
        main="Box Plot of restecg ", 
        col="orange")

```
```{r}

# Boxplot 
boxplot(my_data$thalch, 
        main="Box Plot of thalch", 
        col="orange")

```

```{r}

# Boxplot 
boxplot(my_data$oldpeak, 
        main="Box Plot of oldpeak", 
        col="orange")

```



# Exploring Categorical Data
```{r}

# Count of each num of affected people 
table(my_data$num)

```



```{r}
# Bar plot for count
barplot(table(my_data$num), 
        main="Count of num", 
        col=c("red", "green", "blue"))

```
```{r}
# Bar plot for  count
barplot(table(my_data$sex), 
        main="Count of sex", 
        col=c("red", "green", "blue"))

```
```{r}
# Bar plot for  count
barplot(table(my_data$cp), 
        main="Count of cp", 
        col=c("red", "green", "blue"))

```
```{r}
# Bar plot for  count
barplot(table(my_data$trestbps), 
        main="Count of trestbps", 
        col=c("red", "green", "blue"))

```
```{r}
# Bar plot for count
barplot(table(my_data$slope), 
        main="Count of slope", 
        col=c("red", "green", "blue"))

```


```{r}
# Bar plot for Species count
barplot(table(my_data$thal), 
        main="Count of thal", 
        col=c("red", "green", "blue"))

```





# Multivariate Exploration
# Scatter Plot (Numeric vs. Numeric)
```{r}
# Scatter plot: age vs trestbps
plot(my_data$age, my_data$trestbps, 
     main="Scatter Plot of Age vs trestbps",
     xlab="age", ylab="trestbps",
     col=my_data$num, pch=20)
legend("bottomright", legend=levels(my_data$num), col=1:4, pch=19)

```
# Multivariate Exploration
# Scatter Plot (Numeric vs. Numeric)
```{r}
# Scatter plot: thalch vs chol
plot(my_data$thalch, my_data$chol, 
     main="Scatter Plot of thalch vs chol",
     xlab="thalch", ylab="chol",
     col=my_data$num, pch=20)
legend("topleft", legend=levels(my_data$num), col=1:4, pch=20)

```

```{r}
# Scatter plot: thalch vs trestbps
plot(my_data$thalch, my_data$trestbps, 
     main="Scatter Plot of thalch vs trestbps",
     xlab="thalch", ylab="trestbps",
     col=my_data$num, pch=20)
legend("bottomleft", legend=levels(my_data$num), col=1:4, pch=20)

```



# Correlation (Numeric vs. Numeric)
```{r}
# Correlation between age and trestbs
cor(my_data$age, my_data$trestbps)

```
```{r}
# Correlation between age and trestbs
cor(my_data$age, my_data$chol)

```
```{r}
# Correlation between chol and trestbs
cor(my_data$chol, my_data$trestbps)

```
```{r}
# Correlation between thalch and trestbs
cor(my_data$thalch, my_data$trestbps)

```

```{r}
# Correlation between age and thalch
cor(my_data$age, my_data$thalch)

```
```{r}
```


```{r}
# Correlation between age and slope
cor(my_data$age, my_data$slope)

```

# Correlation matrix
```{r}

# Select only numeric columns
numeric_cols <- sapply(my_data, is.numeric)
correlation_matrix <- cor(my_data[, numeric_cols], use = "complete.obs")

# Print the correlation matrix
print(correlation_matrix)


```
```{r}
library(corrplot)

# Calculate correlation matrix
numeric_cols <- sapply(my_data, is.numeric)
correlation_matrix <- cor(my_data[, numeric_cols], use = "complete.obs")

# Create heatmap with correlation values
corrplot(correlation_matrix, 
         method = "color",
         type = "upper",
         order = "hclust",
         tl.col = "black",
         tl.srt = 45,
         addCoef.col = "black",
         number.cex = 0.7,
         diag = FALSE)
```


```{r}
library(plotly)

plot_ly(
  x = colnames(correlation_matrix),
  y = colnames(correlation_matrix),
  z = correlation_matrix,
  type = "heatmap",
  colorscale = "Viridis",
  hoverinfo = "x+y+z"
) %>% 
  layout(
    title = "Interactive Correlation Matrix",
    xaxis = list(tickangle = 45),
    margin = list(l = 100, r = 100, b = 150)
  )
```

# Box Plot (Numeric vs. Categorical)
```{r}
# Boxplot: Cholesterol by Heart Disease Status
boxplot(chol ~ num, data = my_data,
        main = "Boxplot of Cholesterol by Heart Disease Status",
        xlab = "Heart Disease",
        ylab = "Cholesterol",
        col = c("skyblue", "tomato"))

```
# Box Plot (Numeric vs. Categorical)
```{r}
# Boxplot: Thalch by Sex
boxplot(thalch ~ sex, data = my_data,
        main = "Boxplot of Max Heart Rate by Sex",
        xlab = "Sex",
        ylab = "Max Heart Rate (thalch)",
        col = c("lightpink", "lightblue"))


```
# Box Plot (Numeric vs. Categorical)
```{r}
# Boxplot: Oldpeak by Chest Pain Type
boxplot(oldpeak ~ cp, data = my_data,
        main = "Boxplot of ST Depression (Oldpeak) by Chest Pain Type",
        xlab = "Chest Pain Type",
        ylab = "Oldpeak",
        col = rainbow(length(unique(my_data$cp))))

```
# Box Plot (Numeric vs. Categorical)
```{r}
# Boxplot: Trestbps by Slope
boxplot(trestbps ~ slope, data = my_data,
        main = "Boxplot of Resting BP by Slope of Peak Exercise ST",
        xlab = "Slope Type",
        ylab = "Resting Blood Pressure",
        col = c("orange", "cyan", "lightgreen"))
```


```{r}
library(ggplot2)
ggplot(my_data, aes(x = as.factor(num), y = chol, fill = as.factor(num))) +
  geom_boxplot() +
  ggtitle("Cholesterol Levels by Heart Disease Class") +
  xlab("Heart Disease Class (num)") +
  ylab("Serum Cholesterol (mg/dl)") +
  theme_minimal()



```



install.packages("ggplot2")
# Pair Plot (Multivariate Visualization)
```{r}

pairs(my_data[, c("age", "trestbps", "chol", "thalch", "oldpeak")],
      col = my_data$num,
      pch = 19,
      main = "Pair Plot of Selected Features Colored by Heart Disease")

```


# Using ggplot2 for Enhanced Visualization
```{r}

library(ggplot2)

my_data$num <- as.factor(my_data$num)


ggplot(my_data, aes(x = age, y = thalch, color = num)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Age vs Max Heart Rate (thalch) by Heart Disease Class") +
  xlab("Age") +
  ylab("Max Heart Rate Achieved (thalch)")


```


# Using ggplot2 for Enhanced Visualization
```{r}



ggplot(my_data, aes(x = trestbps, y = chol, color = num)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Resting Blood Pressure vs Cholesterol by Heart Disease Class") +
  xlab("Resting Blood Pressure (trestbps)") +
  ylab("Serum Cholesterol (chol)")


```


# Using ggplot2 for Enhanced Visualization
```{r}

# Scatter plot with regression line for thalch vs chol
ggplot(my_data, aes(x = thalch, y = chol, color = num)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Maximum Heart Rate Achieved vs Cholesterol by Heart Disease Class") +
  xlab("Maximum Heart Rate Achieved (thalch)") +
  ylab("Serum Cholesterol (chol)") +
  theme_minimal()




```
# Using ggplot2 for Enhanced Visualization
```{r}
# Load necessary libraries
library(caret)

# Select numeric columns for scaling
numeric_cols <- c("age", "trestbps", "chol", "thalch", "oldpeak")

# Apply standardization (z-score normalization)
preProcValues <- preProcess(my_data[, numeric_cols], method = c("center", "scale"))
my_data_scaled <- predict(preProcValues, my_data[, numeric_cols])

# Replace original columns with scaled versions
my_data[, numeric_cols] <- my_data_scaled

# Verify the transformation
summary(my_data[, numeric_cols])
```


```{r}
# Binary encoding for sex (Male=1, Female=0)
my_data$sex <- ifelse(my_data$sex == "Male", 1, 0)

# Ensure fbs is integer (fasting blood sugar > 120 mg/dl)
my_data$fbs <- as.integer(my_data$fbs)

# Convert target variable to factor
my_data$num <- as.factor(my_data$num)

# Identify numeric columns for scaling (excluding already encoded categoricals and target)
numeric_cols <- c("age", "trestbps", "chol", "thalch", "oldpeak", "ca")

# Apply standardization (z-score normalization)
library(caret)
preProcValues <- preProcess(my_data[, numeric_cols], method = c("center", "scale"))

# View the structure of your data
str(my_data)
```




```{r}
# Check the first few rows
head(my_data)
```


```{r}
# Function to detect outliers using IQR method
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  x < lower_bound | x > upper_bound
}

# Check for outliers in numeric columns
outliers_list <- lapply(my_data[, numeric_cols], detect_outliers)
outliers_count <- sapply(outliers_list, sum)
print(outliers_count)
```


```{r}
# Boxplots to visualize outliers
par(mfrow = c(2, 3))
for (col in numeric_cols) {
  boxplot(my_data[[col]], main = paste("Boxplot of", col))
}
par(mfrow = c(1, 1))
```


```{r}
# 1. Robust Scaling
library(caret)
robust_scaler <- preProcess(my_data[, numeric_cols],
                          method = c("center", "scale", "YeoJohnson"))
my_data[, numeric_cols] <- predict(robust_scaler, my_data[, numeric_cols])

# 2. Check for zero-IQR columns
zero_iqr_cols <- sapply(my_data[, numeric_cols], function(x) IQR(x, na.rm = TRUE) == 0)
if(any(zero_iqr_cols)) {
  message("Columns with IQR=0: ", paste(names(zero_iqr_cols)[zero_iqr_cols], collapse = ", "))

  valid_cols <- names(zero_iqr_cols)[!zero_iqr_cols]
}

# 3. Multivariate Outlier Detection (with fallback)
library(MASS)

if(length(valid_cols) > 1) {  # Need at least 2 columns for covariance
  tryCatch({
    mcd <- cov.mcd(my_data[, valid_cols])
    mahalanobis_dist <- mahalanobis(my_data[, valid_cols], 
                                  mcd$center, 
                                  mcd$cov)
    cutoff <- qchisq(0.99, df = length(valid_cols))
    outliers <- mahalanobis_dist > cutoff
    
    # 4. Handle outliers (capping approach)
    my_data[outliers, valid_cols] <- 
      apply(my_data[outliers, valid_cols], 2, 
            function(x) pmin(pmax(x, 
                                 quantile(x, 0.01, na.rm = TRUE)),
                            quantile(x, 0.99, na.rm = TRUE)))
  }, error = function(e) {
    message("MCD failed, using regular Mahalanobis: ", e$message)
    # Fallback to regular Mahalanobis
    mahalanobis_dist <- mahalanobis(my_data[, valid_cols], 
                                  colMeans(my_data[, valid_cols]), 
                                  cov(my_data[, valid_cols]))
    cutoff <- qchisq(0.99, df = length(valid_cols))
    outliers <- mahalanobis_dist > cutoff
  })
} else {
  message("Insufficient valid columns for multivariate outlier detection")
  outliers <- rep(FALSE, nrow(my_data))
}

# 5. Final Verification
summary(my_data[, numeric_cols])
boxplot(my_data[, numeric_cols], main = "Post-Processing Distributions")
```



